Constructing Sailing Match Race Schedules: Round Robin Pairing Lists

### 2015-04-07 paper.tex
Most of Alice's changes

There are some conflicts inherent in the rules. Take criteria~4 and~11, assume we have 4 boats, and
in flight $i$ the last match is the pair of skippers ($x,y$). Criterion~11 dictates that skippers $x$
and $y$ must appear in the next flight, $i+1$, and criterion~4 that neither can be first in the next
flight. This forces skippers $x$ and $y$ to compete in flight $i+1$ as the last match, violating
criterion~1. Therefore, although not explicitly stated, there are no legal schedules with less than 6
boats.

@@

In \eqref{var:timeSlot} the array $\timeSlot$ is defined and in \eqref{var:state}, $\state$ is
defined. If variable $\timeSlot[{i}] = k$ and $k \geq 0$ then this skipper is in a match with
skipper~$k$ at time~$i$. Variable $\state[{i}]$ gives the state of this skipper in flight~$i$. The
cases in \eqref{C1} say that a skipper can be in a match in the first time slot in the flight, or in
the middle of the flight\footnote{\ldots actually, not first and not last.}, or in the last time
slot of the flight. Alternatively, if all time slots in a flight are equal to $-1$ then the skipper
is in a bye, and if all time slots in a flight are equal to $-2$ then the skipper has finished all
his matches.

@@

\paragraph{State transitions:} The $\state$ variables are then used to impose match race criterion~4
(if last in flight $i$ then not first in flight $i+1$), criterion~11 (if last in flight $i$ then not
in a bye in flight $i+1$) and criterion~12 (if flight $i$ is a bye then flight $i$ is not first).
These criteria are imposed by \eqref{C3}, which is a deterministic finite automaton (DFA) constraint
\cite{Pesant04}. The transitions for the DFA are shown pictorially in Figure~\ref{skipper5}: arcs
represent a transition from one state to another. The DFA is encoded as a set of transition objects
$\langle q_{i},\iota,q_{j} \rangle$, where a transition is made from state $q_{i}$ to state $q_{j}$
when encountering input $\iota$. In addition we specify the accepting states, which are all states
except the start state (white) and the bye state (pink).  This constraint also ensures that if a
skipper has finished his matches in flight $i$ (i.e.\ $\END$) then he has finished his matches in
flight $i+1$ (and so on).

Figure~\ref{skipper5} also shows the skipper-oriented variables and constraints corresponding to
skipper 5 in the Table~\ref{tab1} schedule. The $\state$ array is presented as coloured boxes, and
below that is the $\timeSlot$ array. The arrows represent the channeling constraints \eqref{C1}. The
$\state$ array is coloured with green representing $\LAST$, yellow for $\MID$, blue for $\FIRST$,
red for $\LAST$ and (not shown) pink for $\BYE$. The schedule of Table~\ref{tab1} is reproduced with
the matches for skipper 5 emboldened.

@@

\paragraph{Match perspective:} We are now in a position to model criterion~1 (each skipper sails
against every other skipper) and optimisation criterion~10 (minimise boat changes).  In the second
half of Figure~\ref{model:stage1} we present a match perspective of the schedule, using a two
dimensional array of variables $\match$ \eqref{V5}, where $\match[{i,j}]$ is the time slot that
skippers $\sigma[{i}]$ and $\sigma[{j}]$ meet in a match. Only the half above the diagonal is
represented, and the lower half is made up of exactly the same variables, i.e.\ $\match[{i,j}]$ is
\emph{exactly} the same constrained variable as $\match[{j,i}]$. Constraint \eqref{C6} states that
a match between skippers $\sigma[{i}]$ and $\sigma[{j}]$ takes place at time $k$ (i.e.\
$\match[{i,j}] = k$) if and only if  skipper $\sigma[{i}]$'s $k^{th}$ time slot is skipper
$\sigma[{j}]$ and conversely that skipper $\sigma[{j}]$'s $k^{th}$ time slot is skipper
$\sigma[{i}]$). Variables \eqref{V6} and constraints \eqref{C7} then convert this from time slots to
flights, i.e.\ $\modMatch[{i,j}]$ is the flight in which skippers $\sigma[{i}]$ and $\sigma[{j}]$
meet for a match. Finally, constraint \eqref{C8} ensures that each skipper's match occurs in
different flights. Also, since $\modMatch[{i,j}] \equiv \modMatch[{j,i}]$, setting rows to be all
different also ensures that columns are all different.

@@

Having produced a schedule that minimises boat changes, we then minimise criterion~6, imbalance, by
extending the model as in Figure~\ref{model:stage2}.  Assuming the first stage produced a schedule
with $\beta$ boat changes we post this as a hard constraint \eqref{C11} in stage 2. For each skipper
$\sigma$ we introduce a zero/one variable $position[{i,j}]$ \eqref{V9} which is 1 if and only if
skipper $\sigma$ is in a match in time $m \cdot j + i$, i.e.\ in the $i^{th}$ match of the  $j^{th}$
flight \eqref{C12}. Imbalance \eqref{V10} is then computed for each position, as the absolute amount
of over or under expected presence in a given position \eqref{C13}. Finally, we compute the maximum
imbalance over all positions for this skipper (\ref{V11}, \ref{C14}).

@@

The final stage, stage 4, orients skippers within matches, such that they are either port (red) or
starboard (green). This is done to satisfy criteria~2 and~3, 5 and~8. A sketch of the constraint
model is given with a supporting diagram, Figure~\ref{oriented}.  A sequence of zero/one constrained
integer variables, of length $n-1$, is produced for each skipper. These correspond to the sequence
of skippers they meet in matches. For example, taking the schedule in Table~\ref{tab2}, skipper
$\sigma[{2}]$ meets skippers in order: 3,0,1,4,5,6, and skipper $\sigma[{5}]$ in the order
4,6,0,1,3,2.  Therefore, skipper $\sigma[{2}]$'s 4$^{th}$ variable must take a different value from
$\sigma[{5}]$ 5$^{th}$ variable. There are summation constraints on each skipper to enforce
criteria~2 and~3 (equal number of port and starboard matches). Criterion~8 is enforced by posting
constraints such that in a match ($\sigma[{i}],\sigma[{j}]$) where $\abs*{i - j} = 1$, the higher
indexed skipper (lower ranked) is on the port side. A DFA constraint is posted for criterion~5
(restricting sequences of port and starboard assignments).

@@

\paragraph{9 skippers}: Two new schedules are presented in Table~\ref{09}, on the left for 6 boats
and on the right 8 boats. Neither of these schedules appear in the ISAF Umpires' Manual. Both
schedules respect all criteria. For 6 boats there are 8 boat changes (no skipper making more than 1
boat change) and a maximum imbalance of 2. For 8 boats there are again 8 boat changes, no skipper
with more than 1 boat change, and perfect balance (each skipper 4 times on starboard side and 4 on
port side).

@@

It was anticipated that this problem would be easy to model. This naivety was due to ``round
robin'', leading us to believe that we could use some of the models proposed in \cite{trick2002},
\cite{trick2008} and \cite{henz2004}. This naivety slowly dissolved as we addressed more and more of
the ISAF criteria. A number of models we produced were scrapped due to misunderstanding about the
actual problem, i.e.\ a communication problem between the authors. Eventually this was resolved by
presenting the majority of the criteria as a deterministic finite automaton. This became our Rosetta
Stone, with the surprising benefit that it not only improved the teams communications, it was also a
constraint!

@@

There are many redundant variables in our model. One of our first valid models was essentially that
shown in Figure~\ref{skipper5}, and did not use $\modMatch$ and $\match$ variables. Performance was
hopeless, struggling to produce a 7 skipper 6 boat schedule in days. The $\modMatch$ and $\match$
variables were added. This improved domain filtering, and a flattened $\match$ array was used as
decision variables. That is, the decisions where ``when do we schedule this match?''.  At this point
we had not yet produced a 9 skipper 6 boat schedule and we suspected that the combined criteria
might exclude such a schedule. A non-constraint model was developed by the fourth author, primarily
to prove that there was no schedule for 9 skippers with 6 boats. This program used a backtracking
search with decision variables being positions within flights, i.e.\ time slots being assigned
matches. Prior to search, symmetry breaking was performed, and at the top of search the matches in
the first flight were set. A solution was found in seconds! With this experience we added in the
$time$ variables, channeled these into the existing model and used these as decision variables,
i.e.\ the question now was ``what match will we do now?''. We also anchored the matches in the first
flight. With these changes we began to produce schedules in acceptable time. The model was then
built upon, incrementally adding criteria. This took a surprisingly short amount of time, sometimes
minutes of coding to add in a new feature. The model was then enhanced so that it would address
optimisation rather than just satisfaction, again a trivial programming task.

>>>>

There are some conflicts inherent in the rules. Take criteria~4 and~11, assume we have 4 boats, and
in flight $i$ the last match is the pair of skippers ($x,y$). Criterion~11 dictates that skippers $x$
and $y$ must appear in the next flight, $i+1$, and criterion~4 that neither can be first in the next
flight. This forces skippers $x$ and $y$ to compete in flight $i+1$ as the last match, violating
criterion~1. Therefore, although not explicitly stated, there are no legal schedules with fewer than 6
boats.

@@

In \eqref{var:timeSlot} the array $\timeSlot$ is defined and in \eqref{var:state}, $\state$ is
defined. If variable $\timeSlot[{i}] = k$ and $k \geq 0$ then this skipper is in a match with
skipper~$k$ at time~$i$. Variable $\state[{i}]$ gives the state of this skipper in flight~$i$. The
cases in \eqref{C1} state that a skipper can be in a match in the first time slot in the flight, or in
the middle of the flight\footnote{i.e.\ not first and not last.}, or in the last time
slot of the flight. Alternatively, if all time slots in a flight are equal to $-1$ then the skipper
is in a bye, and if all time slots in a flight are equal to $-2$ then the skipper has finished all
his matches.

@@

\paragraph{State transitions:} The $\state$ variables are then used to impose match race criterion~4
(if last in flight $i$ then not first in flight $i+1$), criterion~11 (if last in flight $i$ then not
in a bye in flight $i+1$) and criterion~12 (if flight $i$ is a bye then flight $i+1$ is not first).
These criteria are imposed by \eqref{C3}, which is a deterministic finite automaton (DFA) constraint
\cite{Pesant04}. The transitions for the DFA are shown pictorially in Figure~\ref{skipper5}: arcs
represent a possible transition from one state to another. The DFA is encoded as a set of transition objects
$\langle q_{i},\iota,q_{j} \rangle$, where a transition is made from state $q_{i}$ to state $q_{j}$
when encountering input $\iota$. In addition we specify the accepting states, which are all states
except the start state (white) and the bye state (pink).  This constraint also ensures that if a
skipper has finished his matches in flight $i$ (i.e.\ is in state $\END$) then he will also have
finished his matches in flight $i+1$ (and so on).

Figure~\ref{skipper5} also shows the skipper-oriented variables and constraints corresponding to
skipper 5 in the schedule shown in Table~\ref{tab1}. The $\state$ array is presented as coloured boxes, and
below that is the $\timeSlot$ array. The arrows represent the channeling constraints \eqref{C1}. The
$\state$ array is coloured with green representing $\LAST$, yellow for $\MID$, blue for $\FIRST$,
red for $\LAST$ and (not shown) pink for $\BYE$. The schedule of Table~\ref{tab1} is reproduced with
the matches for skipper 5 emboldened.

@@

\paragraph{Match perspective:} We are now in a position to model criterion~1 (each skipper sails
against every other skipper) and optimisation criterion~10 (minimise boat changes).  In the second
half of Figure~\ref{model:stage1} we present a match perspective of the schedule, using a two
dimensional array of variables $\match$ \eqref{V5}, where $\match[{i,j}]$ is the time slot in which
skippers $\sigma[{i}]$ and $\sigma[{j}]$ meet in a match. Only the half above the diagonal is
represented, and the lower half is made up of exactly the same variables, i.e.\ $\match[{i,j}]$ is
\emph{exactly} the same constrained variable as $\match[{j,i}]$. Constraint \eqref{C6} states that
a match between skippers $\sigma[{i}]$ and $\sigma[{j}]$ takes place at time $k$ (i.e.\
$\match[{i,j}] = k$) if and only if  skipper $\sigma[{i}]$'s $k^{th}$ time slot is skipper
$\sigma[{j}]$ and conversely that skipper $\sigma[{j}]$'s $k^{th}$ time slot is skipper
$\sigma[{i}]$). Variables \eqref{V6} and constraints \eqref{C7} then convert this from time slots to
flights, i.e.\ $\modMatch[{i,j}]$ is the flight in which skippers $\sigma[{i}]$ and $\sigma[{j}]$
meet for a match. Finally, constraint \eqref{C8} ensures that each skipper's match occurs in
different flights. Also, since $\modMatch[{i,j}] \equiv \modMatch[{j,i}]$, setting rows to be all
different also ensures that columns are all different.

@@

Having produced a schedule that minimises boat changes, we then minimise imbalance (i.e.\ apply
criterion~6) by extending the model as in Figure~\ref{model:stage2}.  Assuming the first stage
produces a schedule with $\beta$ boat changes we post this as a hard constraint \eqref{C11} in stage
2. For each skipper $\sigma$ we introduce a zero/one variable $position[{i,j}]$ \eqref{V9} which is
1 if and only if skipper $\sigma$ is in a match in time $m \cdot j + i$, i.e.\ in the $i^{th}$ match
of the  $j^{th}$ flight \eqref{C12}. Imbalance \eqref{V10} is then computed for each position, as
the absolute amount over or under the expected presence in a given position \eqref{C13}. Finally, we
compute the maximum imbalance over all positions for this skipper (\ref{V11}, \ref{C14}).

@@

The final stage, stage 4, orients skippers within matches, such that they are either port (red) or
starboard (green). This is done to satisfy criteria~2 and~3, 5 and~8. A sketch of the constraint
model is given with a supporting diagram, Figure~\ref{oriented}.  A sequence of zero/one constrained
integer variables, of length $n-1$, is produced for each skipper. These correspond to the sequence
of skippers they meet in matches. For example, taking the schedule in Table~\ref{tab2}, skipper
$\sigma[{2}]$ meets skippers in order: 3,0,1,4,5,6, and skipper $\sigma[{5}]$ in the order
4,6,0,1,3,2.  Therefore, skipper $\sigma[{2}]$'s 4$^{th}$ variable must take a different value from
$\sigma[{5}]$'s 5$^{th}$ variable. There are summation constraints on each skipper to enforce
criteria~2 and~3 (equal number of port and starboard matches). Criterion~8 is enforced by posting
constraints such that in a match ($\sigma[{i}],\sigma[{j}]$) where $\abs*{i - j} = 1$, the higher
indexed skipper (lower ranked) is on the port side. A DFA constraint is posted for criterion~5
(restricting sequences of port and starboard assignments).

@@

\paragraph{9 skippers}: Two new schedules are presented in Table~\ref{09}, on the left for 6 boats
and on the right for 8 boats. Neither of these schedules appear in the ISAF Umpires' Manual. Both
schedules respect all criteria. For 6 boats there are 8 boat changes (no skipper making more than 1
boat change) and a maximum imbalance of 2. For 8 boats there are again 8 boat changes, no skipper
with more than 1 boat change, and perfect balance (each skipper 4 times on starboard side and 4 on
port side).

@@

It was anticipated that this problem would be easy to model. This naivety was due to the use of the
term ``round robin'' for these schedules, leading us to believe that we could use some of the models
proposed in \cite{trick2002}, \cite{trick2008} and \cite{henz2004}. This naivety slowly dissolved as
we addressed more and more of the ISAF criteria. A number of models we produced were scrapped due to
misunderstanding about the actual problem, i.e.\ a communication problem between the authors.
Eventually this was resolved by presenting the majority of the criteria as a deterministic finite
automaton. This became our Rosetta Stone, with the surprising benefit that it not only improved the
teams communications, it was also a constraint!

@@

There are many redundant variables in our model. One of our first valid models was essentially that
shown in Figure~\ref{skipper5}, and did not use $\modMatch$ and $\match$ variables. Performance was
hopeless, struggling to produce a 7 skipper 6 boat schedule in days. The $\modMatch$ and $\match$
variables were added. This improved domain filtering, and a flattened $\match$ array was used as
decision variables. That is, the decisions where ``when do we schedule this match?''.  At this point
we had not yet produced a 9 skipper 6 boat schedule and we suspected that the combined criteria
might exclude such a schedule. A non-constraint model was developed by the fourth author, primarily
to attempt to prove that there was no schedule for 9 skippers with 6 boats. This program used a
backtracking search with decision variables being positions within flights, i.e.\ time slots being
assigned to matches. At the top of search, the matches in the first flight were set, and symmetry
breaking was used to reduce the number of legal second flights. A solution was found in seconds!
With this experience we added in the $\timeVar$ variables, channeled these into the existing model
and used these as decision variables, i.e.\ the question now was ``what match will we do now?''. We
also anchored the matches in the first flight. With these changes we began to produce schedules in
acceptable time. The model was then built upon, incrementally adding criteria. This took a
surprisingly short amount of time, sometimes minutes of coding to add in a new feature. The model
was then enhanced so that it would address optimisation rather than just satisfaction, again a
trivial programming task.
